---
title: "Introduction to `MortalityLaws`"
author: "Marius D. Pascariu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Intro}
  %\VignetteEncoding{UTF-8}
bibliography: Mlaw_References.bib
biblio-style: "apalike"
link-citations: true
---

```{r setup, include=FALSE}
library(MortalityLaws)
library(knitr)
opts_chunk$set(collapse = TRUE)
```

# Package structure

**MortalityLaws** is an R package that provides tools for fitting and analyzing a wide
range of parametric mortality models. It exploits various optimisation methods to
handle complex models efficiently. The package can construct full and abridged life
tables from different inputs and download data directly from the Human Mortality
Database (HMD).

The core functions include:

- `MortalityLaw` : fit parametric mortality models.
- `LifeTable` : build life tables.
- `LawTable` : generate life tables from a fitted mortality law.
- `convertFx` : convert between mortality functions (e.g., `mx` to `qx`).
- `ReadHMD` : download HMD data.

Support functions such as `availableLaws`, `availableLF`, and `availableHMD` return
information about the implemented mortality laws, loss functions, and HMD data
availability, respectively. Standard generics (`predict`, `plot`, `summary`, `fitted`,
`residuals`) work on `MortalityLaw` objects. `LifeTable` objects have a `print`
method, and the life table itself is a plain `data.frame` stored in the `$lt`
component.

A small built-in dataset (`ahmd`) is provided for testing. All functions are documented
in the usual R way; after loading the package with `library(MortalityLaws)`, you can
type `?MortalityLaw` to see its help file.

# Downloading HMD data

The `ReadHMD` function downloads data from the Human Mortality Database
[@university_of_california_berkeley_human_2016]. To use it, you need a valid HMD
account.

```{r LoadPackages, message=FALSE}
library(MortalityLaws)
```

```{r ReadHMD, eval=FALSE}
# Download Swedish death counts (ages 0–110, years 1751–2014)
HMD_Dx <- ReadHMD(
  what      = "Dx",
  countries = "SWE",
  interval  = "1x1",
  username  = "user@email.com",
  password  = "password",
  save      = FALSE
)
```

You can similarly download:

- `births` : birth records
- `Ex` : exposure-to-risk
- `Dx_lexis`, `Ex_lexis` : deaths and exposures by Lexis triangles
- `population` : population size
- `mx` : death rates
- `LT_f`, `LT_m`, `LT_t` : life tables by sex
- `e0` : life expectancy at birth
- `mxc` : cohort death rates
- `Exc` : cohort exposures

for over 40 countries and regions in various formats (`1x1`, `1x5`, `5x1`, `5x5`).

# Model fitting and diagnosis

Once you have data, fitting a model is straightforward. Below we fit a
Heligman–Pollard [@heligman1980] model to the bundled `ahmd` data (England & Wales,
1950) using the squared log-ratio loss (loss function `"LF2"`).

```{r}
year     <- 1950
ages     <- 0:100
deaths   <- ahmd$Dx[paste(ages), paste(year)]
exposure <- ahmd$Ex[paste(ages), paste(year)]

fit <- MortalityLaw(
  x          = ages,
  Dx         = deaths,
  Ex         = exposure,
  law        = "HP",
  opt.method = "LF2"
)
```

## Examining the fitted model

```{r}
ls(fit)   # components of the fitted object
```

A summary of the fit gives parameter estimates and goodness‑of‑fit statistics:

```{r}
summary(fit)
```

Visual inspection is also easy:

```{r, fig.align='center', out.width='80%', fig.width=9}
plot(fit)
```

The default plot shows observed and fitted age-specific death rates (or probabilities),
the residuals, and a histogram of the residuals.

## Fitting on a subset of ages

Sometimes you may want to fit the model only to a certain age range while still
plotting over the full range. Use `fit.this.x` to specify the ages used during
optimisation:

```{r, fig.align='center', out.width='80%', fig.width=9}
fit.subset <- MortalityLaw(
  x          = ages,
  Dx         = deaths,
  Ex         = exposure,
  law        = "HP",
  opt.method = "LF2",
  fit.this.x = 0:65
)
plot(fit.subset)
```

The grey area indicates the ages that were actually used in the fit.

## Age scaling

Some mortality laws model only a portion of the lifespan, and their exponential or
power terms can become numerically unstable when evaluated at large ages. Such laws are
flagged with `SCALE_X = TRUE` in `availableLaws()`, and `MortalityLaw` rescales the age
vector automatically before fitting:

$$x_{fit} = x - \min(x) + 1,$$

so that the youngest age in the fitting range becomes 1. No user argument is involved:
the flag travels with the law definition. The same transformation is applied
consistently when the fit is evaluated, including in `predict()` and `LawTable()`, so
extrapolated ages and life tables are built on the same age scale as the fit.

Model parameters therefore refer to the scaled ages. They are not back-transformed to
the original age scale, so they are not directly comparable with the parameterisation
of the unscaled law. The fitted and predicted curves themselves are always drawn and
returned on the original age scale.

```{r, eval=FALSE}
# Which laws are fitted on a rescaled age vector?
A <- availableLaws()$table
A[as.logical(A$SCALE_X), c("NAME", "CODE")]
```

# Mortality laws in the package

The table below lists the mortality laws implemented in the package
(`availableLaws()` returns the same list dynamically). The **Code** column shows the
string used in the `law` argument of `MortalityLaw`.

| Mortality Law | Predictor | Code |
|---------------|-----------|------|
| De Moivre | $\mu(x) = \frac{1}{N - x}$ | `demoivre` |
| Gompertz | $\mu(x) = A e^{Bx}$ | `gompertz` |
| Gompertz (reparam.) | $\mu(x) = \frac{1}{\sigma} \exp\left\{\frac{x-M}{\sigma}\right\}$ | `gompertz0` |
| Inverse-Gompertz | $\mu(x) = \frac{\frac{1}{\sigma}\exp\left\{-\frac{x-M}{\sigma}\right\}}{\exp\left\{e^{-(x-M)/\sigma}\right\}-1}$ | `invgompertz` |
| Makeham | $\mu(x) = A e^{Bx} + C$ | `makeham` |
| Makeham (reparam.) | $\mu(x) = \frac{1}{\sigma} \exp\left\{\frac{x-M}{\sigma}\right\} + C$ | `makeham0` |
| Opperman | $\mu(x) = \frac{A}{\sqrt{x}} - B + C \sqrt{x}$ | `opperman` |
| Thiele | $\mu(x) = A_1 e^{-B_1 x} + A_2 e^{-\frac12 B_2 (x-C)^2} + A_3 e^{B_3 x}$ | `thiele` |
| Wittstein | $q(x) = \frac{1}{B} A^{-(B x)^N} + A^{-(M - x)^N}$ | `wittstein` |
| Perks | $\mu(x) = \frac{A + B C^x}{1 + D C^x}$ | `perks` |
| Weibull | $\mu(x) = \frac{1}{\sigma} \left( \frac{x}{M} \right)^{\frac{M}{\sigma}-1}$ | `weibull` |
| Inverse‑Weibull | $\mu(x) = \frac{\frac{1}{\sigma} \left( \frac{x}{M} \right)^{-\frac{M}{\sigma}-1}}{\exp\left\{\left( \frac{x}{M} \right)^{-\frac{M}{\sigma}}\right\}-1}$ | `invweibull` |
| Van der Maen | $\mu(x) = A + Bx + Cx^2 + \frac{I}{N-x}$ | `vandermaen` |
| Van der Maen (simpl.) | $\mu(x) = A + Bx + \frac{I}{N-x}$ | `vandermaen2` |
| Strehler-Mildvan | $\mu(x) = A e^{Bx} \exp\left[-\frac{V}{B}\left(1 - e^{-Bx}\right)\right]$ | `strehler_mildvan` |
| Quadratic | $\mu(x) = A + Bx + Cx^2$ | `quadratic` |
| Beard | $\mu(x) = \frac{A e^{Bx}}{1 + K A e^{Bx}}$ | `beard` |
| Beard-Makeham | $\mu(x) = \frac{A e^{Bx}}{1 + K A e^{Bx}} + C$ | `beard_makeham` |
| Gamma‑Gompertz | $\mu(x) = \frac{A e^{Bx}}{1 + \frac{AG}{B}(e^{Bx}-1)}$ | `ggompertz` |
| Siler | $\mu(x) = A_1 e^{-B_1 x} + A_2 + A_3 e^{B_3 x}$ | `siler` |
| Heligman–Pollard | $\frac{q(x)}{p(x)} = A^{(x+B)^C} + D e^{-E (\ln x - \ln F)^2} + G H^x$ | `HP` |
| Heligman–Pollard (HP2) | $q(x) = A^{(x+B)^C} + D e^{-E (\ln x - \ln F)^2} + \frac{G H^x}{1+G H^x}$ | `HP2` |
| Heligman–Pollard (HP3) | $q(x) = A^{(x+B)^C} + D e^{-E (\ln x - \ln F)^2} + \frac{G H^x}{1+K G H^x}$ | `HP3` |
| Heligman–Pollard (HP4) | $q(x) = A^{(x+B)^C} + D e^{-E (\ln x - \ln F)^2} + \frac{G H^{x^K}}{1 + G H^{x^K}}$ | `HP4` |
| Rogers–Planck | $q(x) = A_0 + A_1 e^{-Ax} + A_2 e^{B(x-U) - e^{-C(x-U)}} + A_3 e^{D x}$ | `rogersplanck` |
| Martinelle | $\mu(x) = \frac{A e^{Bx} + C}{1 + D e^{Bx}} + K e^{Bx}$ | `martinelle` |
| Gompertz-Makeham (GM(1,3)) | $\mu(x) = A_0 + K \exp\left\{B_1 x - B_2 x^2\right\}$ | `makeham_logquad` |
| Gompertz-Makeham (GM(0,3)) | $\mu(x) = K \exp\left\{B_1 x - B_2 x^2\right\}$ | `gompertz_logquad` |
| Carriere 1 | $S(x) = \psi_1 S_1(x) + \psi_2 S_2(x) + \psi_3 S_3(x)$ | `carriere1` |
| Carriere 2 | $S(x) = \psi_1 S_1(x) + \psi_4 S_4(x) + \psi_3 S_3(x)$ | `carriere2` |
| Kostaki | $\frac{q(x)}{p(x)} = A^{(x+B)^C} + D e^{-E_i (\ln x - \ln F)^2} + G H^x$ | `kostaki` |
| Kannisto | $\mu(x) = \frac{A e^{Bx}}{1 + A e^{Bx}}$ | `kannisto` |
| Kannisto-Makeham | $\mu(x) = \frac{A e^{Bx}}{1 + A e^{Bx}} + C$ | `kannisto_makeham` |

The `opperman` formula is evaluated internally at age $x + 1$. The `weibull`
hazard is not defined at birth, so age 0 is reported as missing and takes no
part in the fit.

Use `availableLaws()` in R to see this list dynamically:

```{r, eval=FALSE, warning=FALSE}
availableLaws()
```

# Loss functions in the package

A parametric mortality model is fitted by optimising a loss function (e.g., a negative
log-likelihood or a sum of squared errors). MortalityLaws provides eight loss functions
that can emphasise different portions of the mortality curve. Use `availableLF()` to
view their descriptions:

```{r, message=FALSE, warning=FALSE}
availableLF()
```

# Custom mortality laws

You are not limited to the built-in laws: you can define your own hazard function and
pass it via `custom.law`. As an example, we fit a reparametrised Gompertz model in terms
of the modal age at death $M$ [@missov2015]:

\[
\mu_x = \beta e^{\beta (x - M)}.
\]

First, define the hazard function. The function must return a list containing at least
the hazard vector `hx`.

```{r}
my_gompertz <- function(x, par = c(b = 0.13, M = 45)){
  hx <- with(as.list(par), b * exp(b * (x - M)))
  return(as.list(environment()))   # must return a list
}
```

Select data and fit:

```{r}
year     <- 1950
ages     <- 45:85
deaths   <- ahmd$Dx[paste(ages), paste(year)]
exposure <- ahmd$Ex[paste(ages), paste(year)]
```

```{r, warning=FALSE, results='hide'}
my_model <- MortalityLaw(
  x          = ages,
  Dx         = deaths,
  Ex         = exposure,
  custom.law = my_gompertz
)
```

```{r}
summary(my_model)
```

```{r}
plot(my_model)
```

# Full life tables

The `LifeTable` function can build complete or abridged life tables from several types
of input:

- Death counts and exposures (`Dx`, `Ex`)
- Age‑specific death rates (`mx`)
- Death probabilities (`qx`)
- Survivorship (`lx`)
- Death distribution (`dx`)

Only one of these inputs needs to be specified; the rest are ignored if present.

```{r}
y  <- 1900
x  <- as.numeric(rownames(ahmd$mx))
Dx <- ahmd$Dx[, paste(y)]
Ex <- ahmd$Ex[, paste(y)]

LT1 <- LifeTable(x, Dx = Dx, Ex = Ex)   # primary input
LT2 <- LifeTable(x, mx = LT1$lt$mx)     # from mx
LT3 <- LifeTable(x, qx = LT1$lt$qx)     # from qx
LT4 <- LifeTable(x, lx = LT1$lt$lx)     # from lx
LT5 <- LifeTable(x, dx = LT1$lt$dx)     # from dx

LT1
```

```{r}
ls(LT1)   # components of the life table object
```

# Abridged life tables

Abridged life tables use wider age intervals (e.g., 0, 1–4, 5–9, ...). The same
`LifeTable` function works for any abridged age structure.

```{r}
x  <- c(0, 1, seq(5, 110, by = 5))
mx <- c(.053, .005, .001, .0012, .0018, .002, .003, .004, 
       .004, .005, .006, .0093, .0129, .019, .031, .049, 
       .084, .129, .180, .2354, .3085, .390, .478, .551)

lt <- LifeTable(x, mx = mx, sex = "female")
```

```{r}
lt
```

# Citation

```r
citation(package = "MortalityLaws")
```

# References