---
title: "Life tables and how `MortalityLaws` builds them"
subtitle: "From deaths and rates to survivorship, person-years and life expectancy"
author: "Marius D. Pascariu"
date: "2026-10-05"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    number_sections: false
    fig_width: 7
    fig_height: 4.2
    fig_align: center
    code_folding: hide
bibliography: Mlaw_References.bib
biblio-style: apalike
link-citations: true
vignette: >
  %\VignetteIndexEntry{Life tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(MortalityLaws)
```

A death rate is a fact about one year of one population. A life table turns a
whole schedule of such facts into a statement about lives: how many of 100,000
newborns are still alive at each age, how many die at each age, and how many
years remain on average at each age. That translation is mechanical once the
conventions are fixed. This article is about the conventions, because that is
where all the choices live.

`MortalityLaws` does the translation in one function, `LifeTable()`. Hand it
deaths and exposures, or rates, or probabilities, or a survivorship curve, or a
distribution of deaths, or a curve of life expectancy, and it returns the full
ten-column table. Three points along the way ask for a decision: how rates
become probabilities (the $a_x$ convention), where the table closes (the open
age group), and what to do when the data stop before mortality does (closing
and extending). We take those in order, and finish with the two convenience
wrappers, `convertFx()` for one-off conversions and `LawTable()` for tables
that come from a parametric law rather than from data.

The running example is `ahmd`, the bundled England and Wales female data set:
deaths, exposures and death rates at single ages 0 to 110 for the years 1850,
1900, 1950 and 2010. The rate series thins out at the oldest ages (for 1900 the
rows from age 107 on are missing), so the examples below work on ages 0 to 105,
where every year is populated.

# What a life table is

A life table follows one synthetic cohort of $l_0$ newborns through life and
records, for every age interval $[x, x+n)$, what happens to them. The cohort is
synthetic in a precise sense: at every age it is exposed to the mortality that
the real population showed at that age in one calendar year. The table is a
summary of that year, not a forecast of anyone's actual life.

`LifeTable()` returns an object of class `"LifeTable"` with three parts: `lt`,
the table itself; `call`, the matched call; and `process_date`, the timestamp.
The `lt` data frame has one row per age interval and ten columns:

| column | symbol | what it holds |
|---|---|---|
| `x.int` | | the interval as a label, e.g. `[0,1)` |
| `x` | $x$ | the age at the start of the interval |
| `mx` | $m_x$ | the death rate in $[x, x+n)$ |
| `qx` | $q_x$ | the probability of dying within the interval |
| `ax` | $a_x$ | the average years lived in the interval by those who die in it |
| `lx` | $l_x$ | the survivors at exact age $x$, on the `lx0` scale |
| `dx` | $d_x$ | the deaths in the interval, on the same scale |
| `Lx` | ${}_nL_x$ | the person-years lived in the interval |
| `Tx` | $T_x$ | the person-years remaining at age $x$ |
| `ex` | $e_x$ | the average years remaining, $e_x = T_x / l_x$ |

The radix `lx0` is 100,000 by default, so `lx`, `dx`, `Lx` and `Tx` are counts
in a hypothetical population of 100,000 newborns. Set `lx0 = 1` and the same
columns read as survivorship probabilities instead. The rates and
probabilities do not depend on the choice.

The input can also be a matrix or a data frame with one column per population
or year. Each column then becomes its own table, stacked into one `lt` with a
leading `LT` column carrying the column name, and `plot()` draws one curve per
table.

# The six ways in

Exactly one input case per call. Supplying two of them is an error, which is
the kind of strictness that pays for itself the first time it fires.

| case | arguments | typical source |
|---|---|---|
| counts | `Dx` and `Ex` | vital statistics over census exposures |
| rates | `mx` | published rate series, e.g. the HMD |
| probabilities | `qx` | published life tables |
| survivorship | `lx` | published life tables |
| deaths | `dx` | published life tables |
| expectancy | `ex` | targets, forecasts, model outputs |

## The count case

Given deaths and exposures, the rate $m_x = D_x / E_x$ is computed for you and
the rest of the table follows.

```{r lt-counts}
x  <- 0:105
Dx <- ahmd$Dx[as.character(x), "1900"]
Ex <- ahmd$Ex[as.character(x), "1900"]

lt_counts <- LifeTable(x = x, Dx = Dx, Ex = Ex)
head(lt_counts$lt, 4)
```

Read the first row: 63,454 deaths in 419,375 person-years of exposure at age 0
give $m_0 = 0.1513$. Of the 100,000 synthetic newborns, 13,694 die before their
first birthday, and the 48.25 years of $e_0$ are what the rest of the schedule
adds up to.

## The rate case

If the rates are already published, skip the division.

```{r lt-rates}
mx <- ahmd$mx[as.character(x), "1900"]

lt_rates <- LifeTable(x = x, mx = mx)
head(lt_rates$lt, 4)
```

The two runs agree to the fourth decimal in $e_0$ (48.25279 against 48.25283)
but not digit for digit, and the reason is worth knowing: the bundled `mx`
column is the published rate series, which is not exactly `Dx / Ex` (the two
differ by 0.016 at age 105, where a handful of deaths sit on very small
exposure). Which of the two you treat as the source of truth is your decision,
not the package's.

## Feeding columns back in

Each of the other four cases takes what a previous table produced. Feed any of
the derived columns back and the same table comes out.

```{r six-inputs}
e0 <- c(
  DxEx = lt_counts$lt$ex[1],
  mx   = LifeTable(x = x, mx = lt_counts$lt$mx)$lt$ex[1],
  qx   = LifeTable(x = x, qx = lt_counts$lt$qx)$lt$ex[1],
  lx   = LifeTable(x = x, lx = lt_counts$lt$lx)$lt$ex[1],
  dx   = LifeTable(x = x, dx = lt_counts$lt$dx)$lt$ex[1],
  ex   = LifeTable(x = x, ex = lt_counts$lt$ex)$lt$ex[1]
)
e0
```

All six agree to within $10^{-6}$ of a year in $e_0$, and the `ex` round trip
is exact to machine precision. The `ex` case is the interesting one, because it
runs the table backwards; the section on [the ex
inverse](#the-ex-inverse-a-table-from-life-expectancies) shows how.

# From rates to probabilities

A rate is not a probability. The rate $m_x = D_x / E_x$ can exceed 1 at old
ages, while $q_x$ is a probability and cannot. Turning one into the other needs
an assumption about where inside the interval the deaths occur, and that
assumption is $a_x$: the average number of years lived in the interval by those
who die in it.

With $a_x$ in hand, the exact interval identity gives

$$
q_x = \frac{n \, m_x}{1 + (n - a_x) \, m_x}
$$

which in code is `qx = n*mx/(1+(n-ax)*mx)`, and its inverse

$$
m_x = \frac{q_x}{a_x q_x + n (1 - q_x)}.
$$

The textbook alternative assumes a constant force of mortality inside the
interval, $\mu = m_x$, which integrates to

$$
q_x = 1 - e^{-n m_x},
$$

in code `1-exp(-n*mx)`, with the inverse $m_x = -\log(1 - q_x)/n$.

Which formula you get depends on the $a_x$ method
(see [Choosing ax](#choosing-ax)). Whenever an $a_x$ is known before the
conversion, either because you supplied one or because the method builds the
whole $a_x$ column first, the exact identity is used, so that the `mx`, `qx`
and `ax` columns describe each other exactly. The exponential form is what
`cfm`, `preston` and `coale_demeny` leave in place: those methods convert under
a constant force of mortality and adjust $a_x$ afterwards.

The two agree closely at ordinary rates and part company at the edges:

```{r bridge}
lt <- lt_rates$lt
q_exact <- (1 * lt$mx) / (1 + (1 - lt$ax) * lt$mx)
q_cfm   <- 1 - exp(-1 * lt$mx)

round(data.frame(
  age     = lt$x[1:3],
  mx      = lt$mx[1:3],
  ax      = lt$ax[1:3],
  qx      = lt$qx[1:3],
  q_exact = q_exact[1:3],
  q_cfm   = q_cfm[1:3]
  ), 4)
```

At age 0 the constant force form gives 0.1404 against the exact 0.1369, and the
gap only widens where the rate is large. This is why the exact identity is the
default: it is the one relation that keeps all three columns consistent.

# How the columns are built

Once $q_x$ is settled the rest of the table is bookkeeping. Survivorship falls
by the deaths, and everything else accumulates person-years:

$$
l_{x+n} = l_x \, (1 - q_x), \qquad l_0 = l_{x0}
$$

$$
d_x = l_x - l_{x+n}
$$

$$
{}_nL_x = n \, l_x - (n - a_x) \, d_x
       = a_x \, l_x + (n - a_x) \, l_{x+n}
$$

$$
T_x = \sum_{t = x,\, x+n,\, \ldots}^{\omega} {}_nL_t
$$

$$
e_x = \frac{T_x}{l_x}
$$

$T_x$ sums the person-years from age $x$ to the open age $\omega$, so it is a
cumulative total and $e_x$ is simply that total per survivor. In code the whole
recursion fits in four lines:

```text
lx = lx0 * c(1, cumprod(1 - qx))     dx = lx - lx_shifted
Lx = n*lx - (n - ax)*dx              Lx[N] = lx[N]/mx[N]  (open row)
Tx = rev(cumsum(rev(Lx)))            ex = Tx/lx,  ex[N] = ax[N]
```

We can check the person-years identity against the table we already built:

```{r recursions}
lt <- lt_rates$lt
Lx_hand <- 1 * lt$lx - (1 - lt$ax) * lt$dx

max(abs(Lx_hand - lt$Lx))
```

The two agree to the last digit. One thing this identity quietly says: $L_x$
depends on $a_x$ as much as $q_x$ does, so a life table is only as good as its
$a_x$ convention.

The rates fed into these recursions are rarely raw. Published tables are
graduated first, smoothed so that the columns move plausibly with age; the
classic actuarial treatment is @forfar1988, and fitting a parametric law is one
modern way to do the same job.

# Choosing ax

$a_x$ is the average number of years lived inside the interval by a person who
dies in it. It fixes how the person-years column ${}_nL_x$ is split between the
survivors and the dying, and through that split it shapes the whole table.
`LifeTable()` offers four methods for it, plus the option of supplying your own
values. They differ only at the edges of the age range: infant ages, wide
intervals, and the open interval. Everywhere else they converge on the midpoint
of the interval.

| `ax` | what it does | when to use it |
|---|---|---|
| `andreev_kingkade` (default) | $a_0$ from the Andreev-Kingkade rule keyed on $m_0$ and sex; $n/2$ on the other one-year intervals; the constant-force value on wider ones; $1/m_x$ on the open row; conversion via the exact identity | reproduces the HMD period life tables (@andreevkingkade2015; @wilmoth2025). Use it unless you have a reason not to. |
| `cfm` | constant force of mortality in every interval: $a_x = n + 1/m_x - n/q_x$ | reproducing tables built on that assumption; quick comparisons |
| `preston` | `cfm`, with the Coale-Demeny West $a_0$ and $a_1$ for the first two intervals, expressed in $m_x$ | following the textbook treatment of infant separation (@preston2001) |
| `coale_demeny` | `cfm`, with the original 1983 $q_0$ rule for the first two intervals (the PAS coefficients) | matching Coale-Demeny regional model tables and PAS output (@coaledemeny1983) |

`cfm` has history: the constant-force value $a_x = n + 1/m_x - n/q_x$ was the
package's default $a_x$ method until v2.8.0, when `andreev_kingkade` took over.
It is still available by name, and it is still the fallback inside the default
rule for intervals wider than a year.

The first two intervals of an abridged table show what separates them. The data
below are a textbook abridged schedule: one-year intervals at 0 and 1, five-year
intervals above.

```{r ax-methods}
x_ab  <- c(0, 1, seq(5, 110, by = 5))
mx_ab <- c(.053, .005, .001, .0012, .0018, .002, .003, .004,
           .004, .005, .006, .0093, .0129, .019, .031, .049,
           .084, .129, .180, .2354, .3085, .390, .478, .551)

LT_ax <- lapply(c("andreev_kingkade", "cfm", "preston", "coale_demeny"), function(a)
  LifeTable(x = x_ab, mx = mx_ab, sex = "female", ax = a))
names(LT_ax) <- c("andreev_kingkade", "cfm", "preston", "coale_demeny")

ax_cmp <- t(sapply(LT_ax, function(L) L$lt$ax[1:2]))
colnames(ax_cmp) <- c("a0", "a1")
round(ax_cmp, 3)
```

`preston` and `coale_demeny` differ by a thousandth of a year here and agree
exactly once $m_0$ reaches 0.107. Both collapse onto `cfm` when `sex` is not
given, because the childhood rule is keyed on the sex of the population. The
life expectancies differ in the second decimal:

```{r ax-methods-e0}
e0_ax <- sapply(LT_ax, function(L) L$lt$ex[1])
round(e0_ax, 2)
```

If you have your own $a_x$ values, pass them instead: a scalar applies
everywhere, a vector must have one value per age. A value an interval cannot
support ($a_x m_x > 1$, more person-years than the interval holds) is replaced
with the implied average $1/m_x$, and the affected ages are reported in a
warning.

# The open age group

Every life table must stop somewhere, and the last row is a convention rather
than a measurement. At the open age $N$ (the `+` in `85+`, or wherever you
choose to close), the package applies the standard reciprocal rule
`ax = ex = 1/mx`:

$$
a_N = e_N = \frac{1}{m_N}, \qquad q_N = 1, \qquad {}_nL_N = \frac{l_N}{m_N}.
$$

In words: everyone still in the table dies in the open interval, the interval
lasts exactly $1/m_N$ years on average, and life expectancy there equals that
duration. (A life table always ends with an assumption; at least this one is
printed in the last row.)

```{r open-age}
lt_ab <- LifeTable(x = x_ab, mx = mx_ab, sex = "female")
tail(lt_ab$lt, 2)
```

Read the last row of the abridged table: $m_{110} = 0.551$, so $a_{110} =
e_{110} = 1/0.551 = 1.81$ years, and the 5,313 survivors at 110 contribute
their $9,642 = 5{,}313 / 0.551$ person-years to $T_{110}$.

The rule assumes the force of mortality is roughly constant above the open age.
That holds when the open interval is old (85+ or 90+) and fails when the data
stop at 75+. The `omega` argument is the clean fix when the data stop early: it
extends the table to a later closing age, filling the new rows by extrapolating
a mortality law. How that extrapolation works is the subject of the next
section.

# Closing and extending the table

Two arguments deal with a tail that the data do not reach, and they attack the
problem from different sides.

`close` stays on the age grid you gave it and replaces the open rate with the
model-implied average force of mortality over the open interval, $1/e_N$ from
the fitted law. `omega` extends the age grid to a later closing age and fills
the added rows with the law's rates. Both are named by a mortality-law code
from `availableLaws()`, both are fitted from `fit_from` (age 60 when the input
reaches 85, the last 20 years of the input otherwise), and when `omega` is set
without `close` the extrapolation defaults to `"kannisto"`, the standard
old-age extension.

The example data stop at 75+:

```{r close-omega}
x5  <- c(0, 1, seq(5, 75, by = 5))
mx5 <- c(.053, .005, .001, .0012, .0018, .002, .003, .004,
         .004, .005, .006, .0093, .0129, .019, .031, .049, .084)

lt_plain <- LifeTable(x = x5, mx = mx5)
lt_close <- LifeTable(x = x5, mx = mx5, close = "kannisto")
lt_ext   <- LifeTable(x = x5, mx = mx5, omega = 110)

c(plain = lt_plain$lt$ex[1], close = lt_close$lt$ex[1], omega = lt_ext$lt$ex[1])
```

The default reciprocal close freezes the age-75 rate of 0.084 forever and hands
back $e_{75} = 1/0.084 = 11.9$ years. The fitted law knows better: it raises
the open rate to 0.129 and drops $e_{75}$ to 7.7 years, which moves $e_0$ from
66.6 to 64.7. Extending the grid to 110 with the same law lands between the two
at 65.2, because the closing assumption there applies five ages later. None of
the three is wrong; they answer slightly different questions, and the arguments
let you say which one you mean.

`close` also accepts any other law code, so the tail can be closed with the
same model you fitted to the whole age range. One caveat lives in `LawTable()`
as well: laws that scale the age vector while fitting (the `SCALE_X` column of
`availableLaws()$table` says which) only produce valid tables from the lower
bound of their fitting range upwards.

# convertFx: one indicator at a time

Often you want one column, not a table. `convertFx()` converts a single
indicator into another and returns a plain vector (or the same matrix you
supplied), tagged with the conversion so that `plot()` can show it.

The grid of possibilities is five inputs by seven outputs:

| from \ to | `mx` | `qx` | `dx` | `lx` | `Lx` | `Tx` | `ex` |
|---|---|---|---|---|---|---|---|
| `mx` | | table | table | table | table | table | table |
| `qx` | table | | table | table | table | table | table |
| `dx` | table | table | | direct | table | table | table |
| `lx` | table | table | direct | | table | table | table |
| `ex` | table | table | table | table | table | table | |

Thirty-five combinations. Only the `dx` to `lx` pair and its reverse are direct
arithmetic (a cumulative sum and a difference); every other pair is answered by
building the full life table in between, so the conversion honours whatever
`ax`, `lx0`, `close` or `omega` you pass through `...`.

```{r convertfx}
ex_1900 <- convertFx(x = x, data = mx, from = "mx", to = "ex")
round(ex_1900[1:3], 2)     # subsetting returns plain numbers

dx_1900 <- convertFx(x = x, data = lt_rates$lt$lx, from = "lx", to = "dx")
max(abs(unclass(dx_1900) - lt_rates$lt$dx))
```

The first conversion is a whole life table under the hood, so its values are
identical to the `ex` column built earlier. The second is arithmetic and
matches the table's own `dx` column to machine precision. `plot.convertFx()`
draws the conversion as it happened, the input indicator in one panel and the
output in the other:

```{r convertfx-plot}
plot(ex_1900)
```

Rates and probabilities go on a log scale in whichever panel they appear, and a
matrix input draws one curve per column.

# LawTable: a table from a law

`LawTable()` takes the coefficients of a mortality law and returns the life
table the law implies. It is the natural next step after a fit with
`MortalityLaw()`, and the only thing to watch is which scale the coefficients
live on.

```{r lawtable}
fit <- MortalityLaw(x = 60:100, mx = mx[x >= 60 & x <= 100], law = "gompertz")
coef(fit)

lt_law <- LawTable(x = 60:100, par = coef(fit), law = "gompertz")
round(head(lt_law$lt[, -1], 3), 4)
```

`par` may be a vector or a matrix with one row of coefficients per table, and
the law code must be one of the 38 in `availableLaws()`. Laws that fit $q_x$
(like the Heligman-Pollard family) feed probabilities to `LifeTable()`, the
rest feed rates. The Gompertz above is one of the `SCALE_X` laws: it rescales
the age vector during fitting, so its coefficients refer to ages measured from
the start of the fitting range. Here the fit starts at 60 and the table starts
at 60, and the law puts $e_{60}$ at 13.79 years against the 13.87 of the data
table at the same age. Try the same coefficients at age 20 and the result would
be quietly wrong; keep the lower bound of `x` at the lower bound of the fit.

# The ex inverse: a table from life expectancies

The `ex` input answers the question in reverse: given a curve of remaining life
expectancy, what life table produced it? The method is the standard textbook
inversion (Preston, Heuveline and Guillot 2001, chapter 3, @preston2001): it
combines the interval identity ${}_nL_x = a_x l_x + (n - a_x) l_{x+n}$ with the
$T_x$ recurrence, and both are relations the forward table is built on. The
$a_x$ values it uses are the forward table's own rules: the Andreev-Kingkade
$a_0$ and the HMD protocol basis under the default
(@andreevkingkade2015; @wilmoth2025), the Coale-Demeny variant when you ask for
it (@coaledemeny1983). The recovery rests on one exact identity that holds for
any $a_x$:

$$
e_x \, l_x - e_{x+n} \, l_{x+n} = {}_nL_x = a_x \, l_x + (n - a_x) \, l_{x+n}.
$$

Divide by $l_x$ and the survivorship ratio of each closed interval falls out:

$$
r = \frac{l_{x+n}}{l_x} = \frac{e_x - a_x}{e_{x+n} + n - a_x}
$$

which in code is:

```text
r = (ex - ax)/(ex_n + n - ax)
```

Then $q_x = 1 - r$ and
$m_x = q_x / (a_x q_x + n (1 - q_x))$ follow, and the open interval closes by
the usual rule, $a_N = e_N$ and $m_N = 1/e_N$.

There is one catch, and it is a real one: $e_x$ alone does not pin down $a_x$.
Pass `ax` as a number and the inversion is exact in one pass, because the ratios
come straight from the formula above. Leave it out and it turns into a
fixed-point problem: the default rule makes $a_x$ a function of the rate being
recovered, so each interval must be solved for the ratio $r$ at which the $a_x$
method in force reproduces it, that is, the root of

$$
\frac{e_x - a(r)}{e_{x+n} + n - a(r)} - r = 0
$$

by bisection (120 iterations, on $r \in (0, 1)$). The root is unique: the left
side is continuous and strictly decreasing in $r$ for every method the package
offers. The recovered table is then exactly the one the same $e_x$ curve
describes under that method.

```{r ex-inverse}
lt_back <- LifeTable(x = x, ex = lt_rates$lt$ex)

round(data.frame(
  age    = x[1:3],
  ex_in  = lt_rates$lt$ex[1:3],
  ex_out = lt_back$lt$ex[1:3],
  mx_out = lt_back$lt$mx[1:3]
  ), 5)

max(abs(lt_back$lt$ex - lt_rates$lt$ex))
```

The round trip reproduces the input curve to $10^{-13}$. The reverse is not
guaranteed: not every curve of $e_x$ is a feasible life table. A curve that
rises at young ages is fine (when infant mortality is high, $e_1 > e_0$), but a
value below its own $a_x$, or falling faster than the interval width, gives
death probabilities outside $[0, 1]$ and the function stops with the offending
ages named. A missing value is an error too, never a repaired gap.

# Reading the output

`plot.LifeTable()` draws the four classic panels: survivorship $l(x)$, the
hazard $m(x)$ on a log scale, the death distribution $d(x)$, and life
expectancy $e(x)$.

```{r plot-lifetable, fig.asp = 0.93}
lt_two <- LifeTable(
  x  = x,
  mx = ahmd$mx[as.character(x), c("1900", "2010")]
  )
plot(lt_two)
```

Two curves per panel, one per table. The 1900 curve shows the shape that used
to be normal everywhere: heavy infant mortality, a young-adult hump, then the
Gompertz rise. The 2010 curve has flattened at both ends. One number from each
table captures most of it: $e_0$ was 48.3 years in 1900 and 82.5 in 2010.

`which` selects a single panel and `split` rearranges the layout:

```{r plot-lifetable-one, fig.height = 4.2}
plot(lt_two, which = "hazard")
```

```{r plot-lifetable-split}
plot(lt_two, split = c(1, 4))
```

The `which` values are `"all"` (the default), `"lx"`, `"hazard"`, `"dx"` and
`"ex"`; `split` takes a `c(nrow, ncol)` pair that must hold exactly one slot
per panel. The same four panels are what a `LawTable()` object plots, since it
returns a `"LifeTable"` like any other.

# Session info

```{r session}
sessionInfo()
```

# References
