---
title: "Mortality models and estimation"
author: "Marius D. Pascariu"
date: "2026-10-05"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    number_sections: false
    fig_width: 7
    fig_height: 4.2
    fig_align: center
    code_folding: hide
bibliography: Mlaw_References.bib
biblio-style: apalike
link-citations: true
vignette: >
  %\VignetteIndexEntry{Mortality models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(MortalityLaws)
```

# What a law is

A mortality law is a smooth function of age with a handful of parameters,
meant to capture the age pattern of death in a population. Write $x$ for age at
the start of a one-year interval. There are two standard ways to say how
dangerous that age is, and this package uses both.

* The **force of mortality** (the hazard) $\mu_x$ is the instantaneous rate at
  which a person aged $x$ dies. Think of it as the slope of the survival curve,
  scaled by how many are left to take it.
* The **death probability** $q_x$ is the probability that someone aged $x$ dies
  before reaching $x + 1$. Its complement $p_x = 1 - q_x$ is the chance of
  seeing the year out.

The two are readings of one curve, tied together by the cumulative hazard
$H_x$ and the survivorship $S_x$:

$$
H_x = \int_0^x \mu_t \, dt, \qquad
S_x = e^{-H_x}, \qquad
q_x = 1 - \frac{S_{x+1}}{S_x}, \qquad
p_x = 1 - q_x .
$$

In `MortalityLaws` every law is a function `function(x, par)` returning a list
whose `hx` element holds one value per age. The **FIT** column of the catalogue
below says what `hx` contains: in `mu[x]` laws it is the hazard $\mu_x$, in
`q[x]` laws it is the death probability $q_x$. The fitting engine and
`LawTable()` read `hx` and route it to the right place, so nothing outside the
law definition has to know which kind of law is in play.

One convention is worth knowing before the table: all parameters are strictly
positive. The optimiser works on the log scale (see
[the optimisation procedure](#the-optimisation-procedure)), so where a
published formula wants a negative coefficient, the sign is folded into the
formula itself. Oppermann's middle term is written $-B$, the log-quadratic
families carry $-B_2 x^2$, and so on. Two formula letters are also spelled
differently in R to avoid name clashes: the coefficient $F$ of Thiele's
formula and the hump location $F$ of the Heligman-Pollard family are both the
parameter `F_`, and the Rogers-Planck hump centre $U$ is `U`.

# The catalogue of laws

`availableLaws()` returns exactly the table below, together with the type
legend of the next section. The columns are the code to pass as `law`, the
model name, the formula in the parameter names the law functions use, the
lifespan type, the fitted quantity, whether fitting rescales the ages, the
year the catalogue assigns to the model, and where the formula comes from.

| CODE | NAME | Formula | TYPE | FIT | SCALE_X | YEAR | Reference |
|---|---|---|---|---|---|---|---|
| `demoivre` | De Moivre | $\mu_x = \dfrac{1}{N - x}$ | 6 | `mu[x]` | `FALSE` | 1725 | @demoivre1725 |
| `gompertz` | Gompertz | $\mu_x = A e^{Bx}$ | 3 | `mu[x]` | `TRUE` | 1825 | @gompertz1825 |
| `gompertz0` | Gompertz | $\mu_x = \dfrac{1}{\sigma} \exp\left\{\dfrac{x - M}{\sigma}\right\}$ | 3 | `mu[x]` | `TRUE` | - | @gompertz1825 |
| `invgompertz` | Inverse-Gompertz | $\mu_x = \dfrac{\frac{1}{\sigma} \exp\left\{-\frac{x - M}{\sigma}\right\}}{\exp\left(\exp\left\{-\frac{x - M}{\sigma}\right\}\right) - 1}$ | 2 | `mu[x]` | `TRUE` | - | @tabeau2001 |
| `makeham` | Makeham | $\mu_x = A e^{Bx} + C$ | 3 | `mu[x]` | `TRUE` | 1860 | @makeham1860 |
| `makeham0` | Makeham | $\mu_x = \dfrac{1}{\sigma} \exp\left\{\dfrac{x - M}{\sigma}\right\} + C$ | 3 | `mu[x]` | `TRUE` | - | @makeham1860 |
| `opperman` | Opperman | $\mu_x = \dfrac{A}{\sqrt{x + 1}} - B + C\sqrt{x + 1}$ | 1 | `mu[x]` | `FALSE` | 1870 | @oppermann1870 |
| `thiele` | Thiele | $\mu_x = A e^{-Bx} + C \exp\left\{-\tfrac{1}{2} D (x - E)^2\right\} + F e^{Gx}$ | 6 | `mu[x]` | `FALSE` | 1871 | @thiele1871 |
| `neggompertz` | Negative-Gompertz | $\mu_x = A e^{-Bx}$ | 1 | `mu[x]` | `FALSE` | 1871 | @thiele1871 |
| `wittstein` | Wittstein | $q_x = \dfrac{1}{B} A^{-(Bx)^N} + A^{-(M - x)^N}$ | 6 | `q[x]` | `FALSE` | 1883 | @wittstein1883 |
| `steffensen` | Steffensen | $\mu_x = \dfrac{A + B C^x}{B C^{-x} + 1 + D C^x}$ | 6 | `mu[x]` | `TRUE` | 1930 | @steffensen1930 |
| `perks` | Perks | $\mu_x = \dfrac{A + B C^x}{1 + D C^x}$ | 3 | `mu[x]` | `TRUE` | 1932 | @perks1932 |
| `weibull` | Weibull | $\mu_x = \dfrac{1}{\sigma} \left(\dfrac{x}{M}\right)^{\frac{M}{\sigma} - 1}$ | 1 | `mu[x]` | `FALSE` | 1939 | @weibull1951 |
| `pareto_2` | Pareto-II | $\mu_x = \dfrac{A}{x + C}$ | 1 | `mu[x]` | `FALSE` | 1954 | @lomax1954 |
| `invweibull` | Inverse-Weibull | $\mu_x = \dfrac{\frac{1}{\sigma} \left(\frac{x}{M}\right)^{-\frac{M}{\sigma} - 1}}{\exp\left(\left(\frac{x}{M}\right)^{-\frac{M}{\sigma}}\right) - 1}$ | 2 | `mu[x]` | `TRUE` | - | @weibull1951 |
| `vandermaen` | Van der Maen | $\mu_x = A + Bx + Cx^2 + \dfrac{I}{N - x}$ | 4 | `mu[x]` | `TRUE` | 1943 | @tabeau2001 |
| `vandermaen2` | Van der Maen | $\mu_x = A + Bx + \dfrac{I}{N - x}$ | 5 | `mu[x]` | `TRUE` | 1943 | @tabeau2001 |
| `strehler_mildvan` | Strehler-Mildvan | $\mu_x = A e^{Bx} \exp\left\{-\dfrac{V}{B}\left(1 - e^{-Bx}\right)\right\}$ | 3 | `mu[x]` | `TRUE` | 1960 | @finkelstein2012 |
| `quadratic` | Quadratic | $\mu_x = A + Bx + Cx^2$ | 5 | `mu[x]` | `TRUE` | - | @tabeau2001 |
| `beard` | Beard | $\mu_x = \dfrac{A e^{Bx}}{1 + K A e^{Bx}}$ | 4 | `mu[x]` | `TRUE` | 1971 | @beard1971 |
| `beard_makeham` | Beard-Makeham | $\mu_x = \dfrac{A e^{Bx}}{1 + K A e^{Bx}} + C$ | 4 | `mu[x]` | `TRUE` | 1971 | @beard1971 |
| `ggompertz` | Gamma-Gompertz | $\mu_x = \dfrac{A e^{Bx}}{1 + \frac{AG}{B}\left(e^{Bx} - 1\right)}$ | 4 | `mu[x]` | `TRUE` | 1979 | @vaupel1979 |
| `siler` | Siler | $\mu_x = A e^{-Bx} + C + D e^{E x}$ | 6 | `mu[x]` | `FALSE` | 1979 | @siler1979 |
| `HP` | Heligman-Pollard | $\dfrac{q_x}{p_x} = A^{(x + B)^C} + D \exp\left\{-E \log(x/F)^2\right\} + G H^x$ | 6 | `q[x]` | `FALSE` | 1980 | @heligman1980 |
| `HP2` | Heligman-Pollard | $q_x = A^{(x + B)^C} + D \exp\left\{-E \log(x/F)^2\right\} + \dfrac{G H^x}{1 + G H^x}$ | 6 | `q[x]` | `FALSE` | 1980 | @heligman1980 |
| `HP3` | Heligman-Pollard | $q_x = A^{(x + B)^C} + D \exp\left\{-E \log(x/F)^2\right\} + \dfrac{G H^x}{1 + K G H^x}$ | 6 | `q[x]` | `FALSE` | 1980 | @heligman1980 |
| `HP4` | Heligman-Pollard | $q_x = A^{(x + B)^C} + D \exp\left\{-E \log(x/F)^2\right\} + \dfrac{G H^{x^K}}{1 + G H^{x^K}}$ | 6 | `q[x]` | `FALSE` | 1980 | @heligman1980 |
| `rogersplanck` | Rogers-Planck | $q_x = A_0 + A_1 e^{-Ax} + A_2 \exp\left\{B(x - U) - e^{-C(x - U)}\right\} + A_3 e^{Dx}$ | 6 | `q[x]` | `FALSE` | 1983 | @rogersplanck1983 |
| `martinelle` | Martinelle | $\mu_x = \dfrac{A e^{Bx} + C}{1 + D e^{Bx}} + K e^{Bx}$ | 6 | `mu[x]` | `FALSE` | 1987 | @martinelle1987 |
| `makeham_logquad` | Gompertz-Makeham | $\mu_x = A_0 + K \exp\left\{B_1 x - B_2 x^2\right\}$ | 5 | `mu[x]` | `TRUE` | 1988 | @forfar1988 |
| `gompertz_logquad` | Gompertz-Makeham | $\mu_x = K \exp\left\{B_1 x - B_2 x^2\right\}$ | 5 | `mu[x]` | `TRUE` | 1988 | @forfar1988 |
| `carriere1` | Carriere | $\mu_x = \dfrac{P_1 S^{w}_x \mu^{w}_x + P_2 S^{i}_x \mu^{i}_x + P_3 S^{g}_x \mu^{g}_x}{P_1 S^{w}_x + P_2 S^{i}_x + P_3 S^{g}_x}$, with the Weibull component $\mu^{w}_x = \dfrac{(x/M_1)^{M_1/\sigma_1 - 1}}{\sigma_1}$, $S^{w}_x = \exp\{-(x/M_1)^{M_1/\sigma_1}\}$; the inverse-Weibull component $\mu^{i}_x = \dfrac{(x/M_2)^{-M_2/\sigma_2 - 1}}{\sigma_2\,\bigl(\exp\{(x/M_2)^{-M_2/\sigma_2}\} - 1\bigr)}$, $S^{i}_x = 1 - \exp\{-(x/M_2)^{-M_2/\sigma_2}\}$; and the Gompertz component $\mu^{g}_x = \dfrac{1}{\sigma_3}\exp\left\{\dfrac{x - M_3}{\sigma_3}\right\}$, $S^{g}_x = \exp\{-\exp\{-M_3/\sigma_3\}(\exp\{x/\sigma_3\} - 1)\}$ | 6 | `q[x]` | `TRUE` | 1992 | @carriere1992 |
| `carriere2` | Carriere | $\mu_x = \dfrac{P_1 S^{w}_x \mu^{w}_x + P_2 S^{i}_x \mu^{i}_x + P_3 S^{g}_x \mu^{g}_x}{P_1 S^{w}_x + P_2 S^{i}_x + P_3 S^{g}_x}$, with the Weibull component $\mu^{w}_x = \dfrac{(x/M_1)^{M_1/\sigma_1 - 1}}{\sigma_1}$, $S^{w}_x = \exp\{-(x/M_1)^{M_1/\sigma_1}\}$; the inverse-Gompertz component $\mu^{i}_x = \dfrac{\exp\{-(x - M_2)/\sigma_2\}}{\sigma_2\,\bigl(\exp\{\exp\{-(x - M_2)/\sigma_2\}\} - 1\bigr)}$, $S^{i}_x = \dfrac{1 - \exp\{-\exp\{-(x - M_2)/\sigma_2\}\}}{1 - \exp\{-\exp\{M_2/\sigma_2\}\}}$; and the Gompertz component $\mu^{g}_x = \dfrac{1}{\sigma_3}\exp\left\{\dfrac{x - M_3}{\sigma_3}\right\}$, $S^{g}_x = \exp\{-\exp\{-M_3/\sigma_3\}(\exp\{x/\sigma_3\} - 1)\}$ | 6 | `q[x]` | `TRUE` | 1992 | @carriere1992 |
| `kostaki` | Kostaki | $\dfrac{q_x}{p_x} = A^{(x + B)^C} + D \exp\left\{-\left(E_i \log(x/F)\right)^2\right\} + G H^x$ | 6 | `q[x]` | `FALSE` | 1992 | @kostaki1992 |
| `kannisto` | Kannisto | $\mu_x = \dfrac{A e^{Bx}}{1 + A e^{Bx}}$ | 5 | `mu[x]` | `TRUE` | 1998 | @thatcher1998 |
| `kannisto_makeham` | Kannisto-Makeham | $\mu_x = \dfrac{A e^{Bx}}{1 + A e^{Bx}} + C$ | 5 | `mu[x]` | `TRUE` | 1998 | @thatcher1998 |
| `scholey_shifted_power` | Scholey-Shifted-Power | $\mu_x = A (x + C)^{-B}$ | 1 | `mu[x]` | `FALSE` | 2019 | @scholey2019 |
| `scholey` | Scholey | $\mu_x = A (x + C)^{-B} e^{-Dx}$ | 1 | `mu[x]` | `FALSE` | 2019 | @scholey2019 |

A few notes on the table.

* The formulas mirror the `MODEL` strings of `availableLaws()`, and the
  parameter names are the ones the law functions accept, so they can be
  copied straight into `parS`. Exceptions in spelling: $F$ is `F_` in R (in
  Thiele's formula and in the Heligman-Pollard hump alike), and in `siler` the
  third exponent is the parameter $E$, not Euler's number.
* `opperman` is evaluated at ages shifted by one year, which keeps
  $A/\sqrt{x}$ finite at birth. Its middle term uses the negative branch,
  which is the branch mortality data occupy; the published sign is free.
* `weibull` is undefined at $x = 0$. Age 0 is reported as missing and takes no
  part in any fit.
* `steffensen` is attributed to Steffensen (1930), an attribution that could
  not be verified against the paywalled source. It is the formula this package
  used to ship as `perks`, before the published Perks form was separated out.
* `carriere1` and `carriere2` mix survivorship curves with weights normalised
  to the simplex, so $P_3 = 1 - P_1 - P_2 > 0$. Their `hx` is a numerical
  derivative: the law function builds the mixture survivorship $S_x$, takes
  $H_x = -\log S_x$ and returns the year-to-year increments of that cumulative
  hazard on the age grid, which is the quantity the FIT column labels `q[x]`.
  The analytic hazard is the mixture formula in the table above, the
  derivative of the same $H_x$ [@carriere1992].
* In `kostaki`, $E_i$ is $E_1$ below the cut age $F$ and $E_2$ above it, so the
  accident hump may be asymmetric.
* The two rows dated 1988 are the Gompertz-Makeham graduation forms GM(1,3)
  and GM(0,3); see [the log-quadratic family](#the-log-quadratic-family).
* Where the original publication is not in the bibliography (Van der Maen
  1943, Strehler-Mildvan 1960, and the reparameterised Gompertz and Makeham),
  the Reference column names the closest work the bibliography does carry.

# Lifespan types

The TYPE column says where on the lifespan a law belongs. It is not decoration:
a TYPE 5 law fitted over the whole age range will happily draw nonsense at age
5. `availableLaws()` ships the legend as its second component.

```{r types}
availableLaws()$legend
```

Each type below comes with one representative law, fitted to observed rates to
show the shape. Every figure uses the female population of England and Wales in
2010, from the bundled `ahmd` data, and plots the fitted curve against the
observed rates with `plot(fit, which = "fit")`; the fitted $R^2$ appears in the
figure header. The laws can all be called by name like this, outside of any
fit; each returns its hazard or death probability in `hx`.

**TYPE 1, infant mortality.** The block of `neggompertz`, `weibull`,
`pareto_2`, the two `scholey` laws and `opperman`. The hazard starts high and
falls, steeply at first. `neggompertz` decays at a constant relative rate,
which is a decent story for the post-neonatal months and a poor one for the
first day of life; a power law lets the decay slow down, and
`scholey_shifted_power` fitted to ages 0 to 10 of the 2010 rates tracks the
fall from 0.0041 at birth to 0.00007 at age 10, with $R^2 = 0.999$. One caveat
belongs to the segment rather than to the fit: the infant curve is only fully
identified at day-level resolution, where the shifted-power family of
@scholey2019 is the tool, and since Scholey's daily series is not shipped with
the package, this illustration has to use the single-year rates.

```{r type1, fig.asp = 0.45}
x   <- 0:10
mx  <- ahmd$mx[paste(x), paste(2010)]
fit <- MortalityLaw(x = x, mx = mx, law = "scholey_shifted_power")
plot(fit, which = "fit")
```

**TYPE 2, the accident hump.** `invgompertz` and `invweibull`. The
inverse-Weibull rises from birth to a peak near its location parameter $M$ and
declines after it, which makes it the law to reach for when the hump is
serious. It is also a hard fit on a female schedule, where the hump is a mild
bump: over ages 10 to 35 of the 2010 rates it settles far from the data, with a
negative $R^2$, so the figure shows its sibling instead. The inverse-Gompertz
rises steeply through the young ages and levels off at $1/\sigma$, and over
this window it climbs smoothly past the bump, giving $R^2 = 0.927$.

```{r type2, fig.asp = 0.45}
x   <- 10:35
mx  <- ahmd$mx[paste(x), paste(2010)]
fit <- MortalityLaw(x = x, mx = mx, law = "invgompertz")
plot(fit, which = "fit")
```

**TYPE 3, adult mortality.** `gompertz`, `gompertz0`, `makeham`, `makeham0`,
`perks` and `strehler_mildvan`. Gompertz's claim to fame is that the log
hazard is a straight line in age, which is very nearly true of human adults.
The hazard is unbounded, so these laws are fitted over adult ages rather than
the whole lifespan. Fitted to ages 40 to 80 of the 2010 rates, the line is
straight on the log scale to the eye, and $R^2 = 0.982$.

```{r type3, fig.asp = 0.45}
x   <- 40:80
mx  <- ahmd$mx[paste(x), paste(2010)]
fit <- MortalityLaw(x = x, mx = mx, law = "gompertz")
plot(fit, which = "fit")
```

**TYPE 4, adult and/or old-age mortality.** `vandermaen`, `beard`,
`beard_makeham` and `ggompertz`. The curve rises like a Gompertz through
adulthood and then bends: towards a ceiling, or towards the divergence of
`vandermaen`'s closing term. This is the block to reach for when the old-age
deceleration matters. The figure fits `ggompertz` to ages 40 to 110 of the 2010
rates, and the range matters: over ages 40 to 100 alone the frailty term
collapses to nearly zero and the same law returns a plain Gompertz, while the
longer range gives $R^2 = 0.929$ and a visible bend away from the straight line
above age 100.

```{r type4, fig.asp = 0.45}
x   <- 40:110
mx  <- ahmd$mx[paste(x), paste(2010)]
fit <- MortalityLaw(x = x, mx = mx, law = "ggompertz")
plot(fit, which = "fit")
```

**TYPE 5, old-age mortality.** `vandermaen2`, `quadratic`, `kannisto`,
`kannisto_makeham` and the two log-quadratic laws. These are fitted from the
adult ages up. `kannisto` is the field standard for exactly this job: a
Gompertz-like rise that levels off at a ceiling of one. Fitted to ages 60 to
110 of the 2010 rates it gives $R^2 = 0.924$; the last decade of the range is
thin and noisy, and the ceiling is what keeps the fitted curve steady where the
data stop being informative.

```{r type5, fig.asp = 0.45}
x   <- 60:110
mx  <- ahmd$mx[paste(x), paste(2010)]
fit <- MortalityLaw(x = x, mx = mx, law = "kannisto")
plot(fit, which = "fit")
```

**TYPE 6, the full age range.** `demoivre`, `thiele`, `wittstein`,
`steffensen`, `siler`, the Heligman-Pollard family, `rogersplanck`,
`martinelle`, the two Carriere mixtures and `kostaki`. These are multi-term
curves with an infant part, a hump where the data have one, and a rising
old-age part. Siler's three terms are the cleanest example: immaturity,
background and senescence, added together. Fitted to ages 0 to 100 of the 2010
rates, one curve covers the infant fall, the childhood trough and the adult
rise, with $R^2 = 0.975$.

```{r type6, fig.asp = 0.45}
x   <- 0:100
mx  <- ahmd$mx[paste(x), paste(2010)]
fit <- MortalityLaw(x = x, mx = mx, law = "siler")
plot(fit, which = "fit")
```

# Family biographies

Thirty-eight laws are easier to face in families. What follows is who wrote
what, and what each family is for. Laws whose original is not in the
bibliography lean on @tabeau2001, the standard review of parametric mortality
models, as their umbrella reference.

## The Gompertz-Makeham lineage

@gompertz1825 proposed the oldest law in the catalogue: the force of mortality
rises exponentially with age, $\mu_x = A e^{Bx}$. It is a one-parameter story
about aging, and it is uncannily good at describing adult mortality. The code
`gompertz0` is the same curve written through its mode $M$ and dispersion
$\sigma$, which are readable straight off the fitted line.

@makeham1860 added the age-independent constant, $\mu_x = A e^{Bx} + C$, so
that accidents, infections and other age-free risks have somewhere to live;
`makeham0` is its mode-and-dispersion form. The full memoir of the law is
@makeham1867; see [Bibliographic notes](#bibliographic-notes) for why the
catalogue dates it 1860.

@perks1932 put a logistic denominator on the exponential,
$(A + B C^x)/(1 + D C^x)$, which flattens the rise at the oldest ages. That
four-parameter logistic exists in two spellings in the catalogue, `perks` and
`beard_makeham`, and `beard`, `kannisto` and `kannisto_makeham` are its two-
and three-parameter cases [@beard1971]. They fit identical curves when the
parameter counts match, so pick one and not several.

The Strehler-Mildvan model of vitality decline rounds out the block
[@finkelstein2012]; it predicts a negative correlation between the Gompertz
intercept and slope across populations, a claim worth checking rather than
assuming.

## Siler and the competing risks view

@siler1979 wrote mortality as three independent hazards added together:
immaturity, $A e^{-Bx}$; background, $C$; and senescence, $D e^{Ex}$. The
model came out of animal mortality, where competing risks are the natural
language, and it fits human curves well. Its first term is the declining
hazard Thiele proposed in 1871 [@thiele1871], which the catalogue carries
separately as `neggompertz`. If you want one law with a story for every age
and five parameters, Siler is the usual answer.

## The Heligman-Pollard family

@heligman1980 modelled the odds of dying, $q_x/p_x$, as three terms: the
decline of infancy, a hump shaped like a lognormal density, and the
Gompertz-like rise of old age. It is the standard full-range law in modern
demography, and the fit is genuinely good over the whole lifespan. The
catalogue carries four versions. `HP` is the original; `HP2`, `HP3` and `HP4`
rewrite the old-age term as a logistic, with `HP3` and `HP4` adding a
parameter $K$ that bends the curve further [@heligman1980].

@kostaki1992 extended the family to nine parameters by giving the accident
hump two dispersion parameters, one below and one above the cut age $F$, so
the hump can lean. @martinelle1987 generalised Perks the other way, adding a
linear term $K e^{Bx}$ above the logistic plateau so that very old ages can
keep rising. All of these are high-parameter models and they all want
`opt.method = "LF2"`; see [Fitting](#fitting-a-law-with-mortalitylaw).

## Old age: the Kannisto family

@thatcher1998 studied the force of mortality at ages 80 to 120, and the
logistic they used, $\mu_x = A e^{Bx} / (1 + A e^{Bx})$, is `kannisto`. It
rises like a Gompertz at the younger old ages and levels off at a ceiling of
one, which is exactly the behaviour observed in the sparse, noisy data at the
top of a life table. It is the field standard for closing life tables, and
`LifeTable()` uses it for its `close` and `omega` arguments. Add the constant
$C$ and you get `kannisto_makeham`.

## Carriere mixtures

@carriere1992 took a different tack: instead of one hazard, mix standard
survivorship curves. `carriere1` combines Weibull, inverse-Weibull and
Gompertz components; `carriere2` swaps the middle one for an inverse-Gompertz.
The weights are normalised to the simplex as they are fitted, so the third
weight stays positive. The result is a flexible whole-lifespan curve where
each component has a recognisable job: childhood, hump, old age.

## The infant laws of Scholey

@scholey2019 traced the age trajectory of infant mortality in US register
data and found two regimes: a power law right after birth, then a constant
exponential decline. The product of the two is `scholey`,
$A (x + C)^{-B} e^{-Dx}$; switch the truncation off and you have
`scholey_shifted_power`, $A (x + C)^{-B}$. The family nests the other infant
laws: $D = 0$ gives the shifted power, $B = 1$ with $D = 0$ gives
`pareto_2`, and $B = 0$ gives `neggompertz`. The truncation parameter is only
identifiable on day- or week-level data over the first year of life, which is
a limitation to know before fitting; see
[How to get it wrong](#how-to-get-it-wrong).

## Pareto, Lomax and the power hazards

`pareto_2`, $\mu_x = A/(x + C)$, is the hazard of the Pareto type II, or
Lomax, distribution [@lomax1954]: a shifted power law with the exponent fixed
at one. It is a workhorse for the infancy and childhood years, and it is the
infancy term of the delay-and-compression model of @debeer2016. @vaupel1983
showed the same hazard arises from a gamma-exponential frailty model, so the
smooth curve has a population story behind it.

## The log-quadratic family

`makeham_logquad` and `gompertz_logquad` write the adult component as
$K \exp(B_1 x - B_2 x^2)$, with and without a Makeham constant $A_0$. The
$-B_2 x^2$ term bends the Gompertz straight line down at the oldest ages, so
the hazard decelerates instead of running away. These are the GM(1,3) and
GM(0,3) forms of the Gompertz-Makeham graduation family catalogued by
@forfar1988, which is what the 1988 rows of the table refer to. Whether
mortality really decelerates at the oldest ages is a question with a long
argument attached: @finkelstein2012 works through it in the context of the
Strehler-Mildvan model, and @missov2015 rewrites the Gompertz hazard through
the modal age at death, the parameterisation used by the custom-law example in
the introduction. Note that the sign of $B_2$ is fixed to the decelerating
branch; the accelerating branch cannot be fitted.

## Rogers-Planck

`rogersplanck` is a three-term death probability covering the whole lifespan,
developed as a general schedule for model life tables [@rogersplanck1983]. Its
middle term, $A_2 \exp\{B(x - U) - e^{-C(x - U)}\}$, is a Gompertz-shaped hump
on the log scale around the centre $U$. The companion PAA paper of 1984
[@rogers1984] is a different work and both are kept; see
[Bibliographic notes](#bibliographic-notes).

## Heterogeneity: a note on what a fitted curve describes

Every law in this catalogue describes a population, not a person. Mix people
with different frailties and the aggregate hazard flattens at old ages even if
every individual's hazard runs away: this is the point of @vaupel1979, whose
gamma-frailty Gompertz is exactly the `ggompertz` law in the catalogue. The
dynamics of such mixed populations are worked out in @vaupel1983. So a
decelerating fit is evidence about the curve, and only indirectly about aging.

## Choosing among the families

Pick the lifespan type first, so the law is used where it is defined. Then
pick the family whose terms match the story you want to tell (immaturity,
hump, senescence), and only then worry about the parameter count. A
nine-parameter law will fit any single mortality schedule better than a
two-parameter one; whether it describes anything is a separate question.
@tabeau2001 reviews the field, and @harper1936 remains useful background for
the infant laws, though it maps to no code in the catalogue.

## Bibliographic notes

Two date conflicts run through this documentation, and they are resolved here
once.

* **Makeham.** `availableLaws()` prints 1860 next to the Makeham formula, so
  the catalogue table cites @makeham1860, the short Assurance Magazine note.
  @makeham1867, "On the law of mortality", is the full memoir of the law, and
  that is the work cited in the narrative above.
* **Siler.** `availableLaws()` prints 1979 next to the Siler formula, so the
  catalogue table cites @siler1979, the competing-risk paper in *Ecology*.
  The key `siler1983` is kept in the bibliography but deliberately uncited, because
  what it contains is not verified; if it ever is, it may be cited in the Siler
  note above and nowhere else.

Three smaller points. The catalogue dates the Weibull law 1939; the
bibliography carries the 1951 journal publication [@weibull1951]. The
Rogers-Planck entry follows the 1983 IIASA working paper
[@rogersplanck1983], with @rogers1984 kept as the separate PAA meeting paper.
And the `steffensen` formula is attributed to Steffensen (1930)
[@steffensen1930] on the strength of that attribution alone.

# Fitting a law with MortalityLaw

Everything so far is vocabulary. `MortalityLaw()` is where the data comes in.

```r
MortalityLaw(x, Dx = NULL, Ex = NULL, mx = NULL, qx = NULL,
             law = NULL, opt.method = "LF2", parS = NULL,
             fit.this.x = x, custom.law = NULL, show = FALSE, ...)
```

* `x` is the age vector, one value per age interval, at the start of the
  interval.
* The data is exactly one of three cases: death counts with exposures
  (`Dx` + `Ex`), central death rates (`mx`), or death probabilities (`qx`).
  Give two cases and the engine stops. A matrix or data frame of `Dx` (or
  `mx`, or `qx`) with several columns fits one model per column.
* `law` is a code from [the catalogue](#the-catalogue-of-laws), and
  `custom.law` is your own function; exactly one of the two.
* `opt.method` chooses what "best fit" means. There are eight options,
  discussed below.
* `parS` are starting values for the parameters, as a named, strictly
  positive numeric vector. Leave it alone and the built-in starting values of
  the law are used.
* `fit.this.x` restricts the optimisation to a subset of the ages. The
  fitted values are still returned over all of `x`.
* `show = TRUE` displays a progress bar, which is mostly useful when many
  columns are fitted at once.
* `...` is passed to or from other methods.

The workhorse data set for the examples in these vignettes is the bundled
`ahmd`: death counts, exposures and rates for England and Wales females, ages
0 to 110, for 1850, 1900, 1950 and 2010
[@hmd2026]. Here is a Gompertz fitted to
adult ages in 2010.

```{r fit-gompertz}
year     <- 2010
ages     <- 45:90
deaths   <- ahmd$Dx[paste(ages), paste(year)]
exposure <- ahmd$Ex[paste(ages), paste(year)]

fit <- MortalityLaw(
  x          = ages,
  Dx         = deaths,
  Ex         = exposure,
  law        = "gompertz",
  opt.method = "poissonL"
)
summary(fit)
```

## The eight objectives

What the optimiser minimises is chosen with `opt.method`. Write $\mu$ for the
fitted value and $\nu$ for the observed one: $\nu = D_x/E_x$ when counts and
exposures are supplied, and $\nu = m_x$ or $\nu = q_x$ when rates or
probabilities are. $D_x$ is the death count and $E_x$ the exposure to risk.
Two of the eight are likelihoods, and six are loss functions.

$$
\begin{aligned}
L_{\text{poissonL}}  &= -\bigl[D_x \log \mu - \mu E_x\bigr], \\
L_{\text{binomialL}} &= -\bigl[D_x \log(1 - e^{-\mu}) - (E_x - D_x)\mu\bigr], \\
L_{\text{LF1}} &= \Bigl(1 - \frac{\mu}{\nu}\Bigr)^2, \qquad
L_{\text{LF2}} = \Bigl(\log\frac{\mu}{\nu}\Bigr)^2, \\
L_{\text{LF3}} &= \frac{(\nu - \mu)^2}{\nu}, \qquad
L_{\text{LF4}} = (\nu - \mu)^2, \\
L_{\text{LF5}} &= (\nu - \mu) \log\frac{\nu}{\mu}, \qquad
L_{\text{LF6}} = \lvert \nu - \mu \rvert .
\end{aligned}
$$

The total loss is the sum over the fitted ages; a term that is not finite is
replaced by a penalty of $10^5$, and a fitted hazard larger than one is capped
at one. When rates or probabilities are the input, the observed rates stand in
for the death counts in the two likelihoods. `availableLF()` prints the same
formulas at the console.

```{r availableLF}
availableLF()
```

Choosing is mostly empirical. The two likelihoods are the principled choice
for count data, and `poissonL` works well for most laws. The loss functions
are for when the fit needs help in a particular corner of the curve, and the
high-parameter laws of the Heligman-Pollard family are the reason the option
exists: they fit reliably under `LF2` and less reliably elsewhere, and the
package says so at the console when you try. Test a couple of objectives on
your data before deciding; the results will differ slightly.

# The optimisation procedure

**Starting values.** Every catalogue law carries built-in starting values,
visible by calling the law without `par`.

```{r starting-values}
gompertz(x = 45:90)$par
```

Pass `parS` to supply your own. The names must match the law's parameters
exactly and all values must be positive; the engine validates both before the
optimiser runs. The starting point of the search is `log(parS)` either way.

```{r fit-parS}
fit_parS <- MortalityLaw(
  x          = ages,
  Dx         = deaths,
  Ex         = exposure,
  law        = "gompertz",
  opt.method = "poissonL",
  parS       = c(A = 0.001, B = 0.05)
)
rbind(default = coef(fit), parS = coef(fit_parS))
```

Two very different starting points, one optimum. That is the usual outcome
for these laws, not a guarantee; a fit that lands somewhere else is telling
you the objective has more than one basin.

**The log transform.** Parameters are optimised on the log scale: the
objective calls the law as `fn(x, exp(par))`. This is why all parameters are
strictly positive, why a formula needing $-0.5$ carries the sign in its own
text instead, and why an extreme probe simply underflows rather than
producing a negative hazard.

**The optimiser.** All laws are fitted with `nlminb` and its PORT routines,
with `eval.max` and `iter.max` at 5000. The single exception is `invweibull`,
which uses `optim` with Nelder-Mead. There is no user-facing switch, and the
coefficients returned are always on the original parameter scale.

**Reading the convergence output.** A non-zero convergence code raises a
warning of the form "MortalityLaw: optimisation did not converge (code k)",
and the full optimiser object is kept in `opt.diagnosis` for inspection. Note
that `nlminb` attaches a message such as "relative convergence (4)" even on
success, so a message with `convergence = 0` is routine. `summary()` reports
the iteration count and the convergence status in one line.

**Age scaling.** Laws flagged `SCALE_X = TRUE` are fitted on shifted ages,

$$
x_{\text{fit}} = x - \min(x) + 1,
$$

so that the youngest fitted age becomes 1 and the exponentials stay tame. The
coefficients therefore refer to scaled ages and are not comparable with the
unscaled parameterisation, while the fitted curves are always returned on the
original age scale. The same shift is reapplied whenever the fitted law is
evaluated again, including `predict()` and `LawTable()`. The one trap: for a
scaled law, `LawTable()` is only valid from the lower bound of the fitting
range upwards.

**The fitting window.** `fit.this.x` fits a subset of the ages while keeping
the full fitted curve. This is how you fit an adult law to adult ages without
dropping the rest of the vector.

```{r fit-window}
fit_window <- MortalityLaw(
  x          = ages,
  Dx         = deaths,
  Ex         = exposure,
  law        = "gompertz",
  opt.method = "poissonL",
  fit.this.x = 60:90
)
range(fit_window$input$fit.this.x)
length(fit_window$fitted.values)
```

# Goodness of fit

A finished `MortalityLaw` object carries three kinds of residual (raw,
deviance and Pearson), the `deviance`, the degrees of freedom and
`dispersion`, and `goodness.of.fit` with the log-likelihood, AIC and BIC.

**Information criteria.** For the two likelihood objectives,

$$
\text{logLik} = -L, \qquad
\text{AIC} = 2k - 2\,\text{logLik}, \qquad
\text{BIC} = \log(n)\, k - 2\,\text{logLik},
$$

with $k$ the number of parameters and $n$ the number of fitted ages. All
three are `NaN` for the six loss functions, because a sum of squares is not a
likelihood and pretending otherwise only produces numbers that look like
AICs. One caveat the engine is honest about: the log-likelihood drops
data-only additive constants, so its absolute value differs from what `glm`
reports, while comparisons between fits of the same data are unaffected.

```{r gof-count}
fit$goodness.of.fit
fit$df
```

**Counts or rates.** The deviance and the dispersion depend on the data case.
With counts and exposures, everything follows the Poisson definitions: the
Pearson residual is $(D_x - \mu E_x)/\sqrt{\mu E_x}$, the deviance is the sum
of squared Poisson deviance residuals (the quantity `poissonL` minimises), and

$$
\text{dispersion} = \frac{\sum \text{Pearson}^2}{\text{rdf}},
$$

which is about one for a correctly specified Poisson model and larger when the
data are noisier than Poisson, as single-age death counts usually are. With
rates or probabilities there is no count likelihood, so the residuals are log
residuals, $\log \nu - \log \mu$; the deviance is their sum of squares, and
the dispersion is their mean square.

```{r gof-rate}
fit_rate <- MortalityLaw(
  x          = ages,
  mx         = ahmd$mx[paste(ages), paste(year)],
  law        = "gompertz",
  opt.method = "LF2"
)
fit_rate$goodness.of.fit   # NaN: LF2 is a loss, not a likelihood
c(deviance = fit_rate$deviance, dispersion = fit_rate$dispersion)
```

**Residuals.** Numbers tell you the fit is close; the residual panels tell you
where it is not. `plot(fit)` draws the observed points against the fitted line
plus four diagnostics (residuals against age, against the fitted values, a
normal Q-Q plot and a histogram). The worked example is in the introduction
to the package.

# How to get it wrong

Most of the ways to get it wrong produce output that looks fine. These are
the ones the package tries to warn you about, and one or two it cannot.

* **Fitting `weibull` at birth.** The hazard is undefined at $x = 0$, so age
  0 is dropped from the fit with a warning. Start the fit at age 1 and save
  yourself the surprise.
* **Extrapolating `demoivre`.** The hazard $1/(N - x)$ diverges at $N$, and
  the fit always places $N$ just above the top fitted age. Predicting past the
  fitted range gives a negative hazard, and every `demoivre` fit warns about
  it, naming the fitted $N$.
* **Trusting `scholey` on year data.** The truncation parameter $D$ is only
  identified on day- or week-level ages over the first year of life. On
  single-year ages it collapses to the optimisation boundary and the model
  reduces to `scholey_shifted_power`, with a warning. The fit is still fine;
  it is just a different model than the one you asked for.
* **Ignoring the `kostaki` guard.** If the two hump dispersions drift more
  than a factor of 50 apart, the engine pulls $E_2$ back to $E_1/50$ before
  the hazard is evaluated. This is a deliberate hack to stop an artificial
  jump at the cut age; if the reported coefficients look oddly tidy, this is
  why.
* **Fitting the HP family with a likelihood.** `HP`, `HP2`, `HP3`, `HP4` and
  `kostaki` are reliably estimated under `opt.method = "LF2"` and not under
  the others. The package prints a hint at the console when you try; the hint
  is correct.
* **Plugging scaled coefficients into the unscaled formula.** Laws flagged
  `SCALE_X = TRUE` are calibrated on shifted ages, $x - \min(x) + 1$, so their
  coefficients are not comparable with the unscaled parameterisation and must
  not be read into the formula with ages as they come. The recipe is one line:
  shift the same way before you evaluate. The
  [Age scaling](Intro.html#age-scaling) section of the introduction shows the
  worked demonstration.
* **Reporting AIC from a loss function.** `goodness.of.fit` is `NaN` there on
  purpose. Do not fill it in by hand or compare such numbers across fits.
* **Fitting `perks` and `beard_makeham` to different data sets and comparing
  them.** They are the same four-parameter curve written two ways.

# Custom laws

Anything not in the catalogue can be fitted by handing a function to
`custom.law`. The contract is small.

```r
my_law <- function(x, par) {
  # compute one mortality value per age, on the scale of your data
  list(hx = hx, par = par)      # optionally add Sx = survivorship
}
```

* The function takes the age vector `x` and a parameter vector `par`, and
  returns a list. The `hx` element is required and must hold one value per
  age; `par` is echoed back with the (possibly normalised) parameters, and an
  `Sx` element with the survivorship is optional. The mixture laws of the
  catalogue use `Sx` internally, and so may yours.
* **`hx` must be on the scale of the data you fit.** The engine compares `hx`
  directly with $D_x/E_x$, $m_x$ or $q_x$. A hazard fitted to rate data is
  correct only because $m_x$ is a rate; if you fit probabilities, return
  probabilities. This is what the FIT column means for your own law.
* **Starting values come from `my_law(1)$par`.** The engine evaluates the
  function at `x = 1` and reads `par` out of the returned list, so the
  default value of `par` is where your search starts.
* **Parameters are optimised on the log scale**, so every parameter must be
  positive for the fit to be reachable. Fold any sign into the formula, as the
  built-in laws do.
* **A custom law is always treated as a scaled one.** Ages are shifted by
  $x - \min(x) + 1$ before your function sees them, exactly as for
  `SCALE_X = TRUE` laws, and the fitted curve is reported on the original
  scale.
* Values of `hx` that are non-finite or non-positive are set aside and
  penalised, so a law that misbehaves part of the time drags the fit instead
  of crashing it. Return a clean vector and the optimiser will thank you.
* `LawTable()` takes catalogue codes only; to build life tables from a custom
  law, evaluate it yourself and pass the result to `LifeTable()`.

A worked custom-law fit, with a Gompertz written through the modal age at
death, is in the introduction to the package.

# Session info and references

Everything on this page was produced with:

```{r session}
sessionInfo()
```

## References
