---
title: "MortalityLaws R package"
author: "Marius Pascariu"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
  #rmarkdown::html_vignette 
vignette: >
  %\VignetteIndexEntry{Vignette --- MortalityLaws R package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Warning: This is just a draft and might change at any time 

`MortalityLaws` in an R package which exploits the available optimization methods to provide tools for fitting and analyzing a wide range of mortality models. 

Three main functions are available in the package: `ReadHMD`, `MortalityLaw`, `LifeTable` and `convertFx`. The package provides also generic functions like `summary`, `plot`, `coef`, `fitted.values` and a small data set for testing purposes `ahmd`.


## The structure of the MoralityLaws R package

```{r echo = FALSE, message = FALSE, fig.height = 4.5, fig.width = 7, fig.pos = 'H'}
library(diagram)

names <- c('MortalityLaws\n R package', 
           'ReadHMD()', 'MortalityLaw()', 'LifeTable()', 'convertFx()',
          'summary()', 'coef()','fitted.values()', 'plot()')

M <- matrix(nrow = 9, ncol = 9, byrow = TRUE, 
           data = c(# M hmd, r  m   l  p  s  p
                      0, 0, 0, 0,  0, 0, 0, 0, 0, #H
                      1, 0, 0, 0,  0, 0, 0, 0, 0, #hmd
                      2, 0, 0, 0,  0, 0, 0, 0, 0, #read
                      2, 0, 0, 0,  0, 0, 0, 0, 0, #Ml
                      2, 0, 0, 0,  0, 0, 0, 0, 0, #life
                      0, 0, 3, 0,  0, 0, 0, 0, 0, #summary
                      0, 0, 3, 0,  0, 0, 0, 0, 0, #coef
                      0, 0, 3, 0,  0, 0, 0, 0, 0, #fitted
                      0, 0, 3, 0,  0, 0, 0, 0, 0  #plot
                    ))

par(mar = c(0.1, 0.1, 0.1, 0.1))
plotmat(M, pos = c(1, 4, 4), curve = 0, name = names,
        lwd = 1, box.lwd = 2, cex.txt = 0,
        box.type = "square", box.prop = 0.4, arr.type = "triangle",
        arr.pos = 0.6, shadow.size = 0.01, prefix = "f",
        main = "")
```

All functions are documented in the standard way, which means that once you load the package you can just type for example `?MortalityLaw` to see the help file for `MortalityLaw` function.

## Installation
`MortalityLaws` is a software in development. The repository containing the development version of the package can be found on [GitHub](https://github.com/mpascariu/MortalityLaws) and can be installed by running the following code in the R console:

```{r echo=TRUE, eval=FALSE}
# Make sure you have the most recent version of R and devtools package already installed.
# install.packages("devtools")
devtools::install_github("mpascariu/MortalityLaws")
```

The package is loaded within **R** as follows: 

```{r echo=TRUE, eval=TRUE}
library(MortalityLaws)
```


## Download HMD Data 
Download data form [Human Mortality Database](http://www.mortality.org) using the `ReadHMD` function:
 
```{r echo=TRUE, eval=FALSE}
# Download HMD data - death counts
HMD_Dx <- ReadHMD(what = "Dx",
                  countries = "SWE",  # HMD country code for Sweden
                  interval = "1x1",            # 1 year - 1 age data
                  username = "user@email.com", # here add your HMD username 
                  password = "password",       # here add your password account
                  save = FALSE)                # Do you what to save the data outside R
```
Here we have downloaded all the registered death counts in Sweden from 1751 until 2014. In the same way one can download the following records: birth counts, population size, lexis triangles, exposures, life tables, life expectancy at birth and death rates for over 38 countries and regions in different formats.


Download life tables all Swedish life tables for female population and save them in an external object:
```{r echo=TRUE, eval=FALSE}
HMD_Ex <- ReadHMD(what = "LT_f", 
                  countries = "SWE", 
                  interval = "1x1", 
                  username = "user@email.com", 
                  password = "password", 
                  save = TRUE)
```


\subsection{Model fitting and diagnosis}
Once we have data from HMD or other sources we can start analyzing it.
For example, let's fit a Heligman-Pollard model under a Poisson setting which is already implemented as one of the standard models in the package\footnote{Check the Appendix to see all the implemented mortality laws.}. We have to use the `MortalityLaw` function in this regard.
In the examples provided below we will use the test data provided in the package `ahmd`.

```{r fitmodel, echo=TRUE, eval=TRUE}
# Select 1 year of data, an age-range and sex
year = 1950
ages = 0:100
deaths <- ahmd$Dx[paste(ages), paste(year)]
exposure <- ahmd$Nx[paste(ages), paste(year)]

fit <- MortalityLaw(x = ages,
                    Dx = deaths, # vector with death counts
                    Ex = exposure, # vector containing exposures
                    law = "HP",
                    how = "poissonL",
                    fit.this.x = 0:65)

ls(fit) # inspect the output object
```


A summary can be obtained using the `summary` function.
```{r echo=TRUE, eval=TRUE}
summary(fit)
```

The standard plot helps us to investigate visually the goodness of fit.
```{r echo = TRUE, eval = TRUE, fig.cap='Heligman-Pollard model fitted using the MortalityLaw function', fig.height = 3, fig.width = 7, fig.pos='H'}
plot(fit)
```

The gray area on the plot showing the fitted value indicates the age range used in fitting the model. This is specified in the model design with the help of \texttt{fit.this.x} argument, which can be  adjusted accordingly.\\

We can use the fitted values to compute life tables. This can be achieved using the \texttt{lifetable} function.
```{r echo=TRUE, eval=TRUE}
lt <- LifeTable(x = ages, mx = fitted.values(fit))
ls(lt)
head(lt$lt)
```

Now let's fit a mortality law that is not defined in the package, say a reparametrize version of Gompertz in terms of modal age at death
<!-- \citep{missov2015gompertz}, -->

\begin{equation}
\mu_x = \beta e^{\beta (x-M)}.
\end{equation}

We have to define a function containing the desired hazard function and then using the `custom.law` argument it can be used in the `MortalityLaw` function.

```{r gompertz2, echo=TRUE, eval=TRUE, fig.cap='Gompertz model fitted using the MortalityLaw function', fig.height = 3, fig.width = 7, fig.pos='htb'}
# Select the data that we want to fit
year = 2010
ages = 35:75
deaths <- ahmd$Dx[paste(ages), paste(year)]
exposure <- ahmd$Nx[paste(ages), paste(year)]

# Here we define a function for our new model and provide start parameters
my_gompertz <- function(x, par = c(b = 0.13, M = 45)){
  hx  <- with(as.list(par), b*exp(b*(x - M)) )
  return(as.list(environment())) # return everything inside this function
}

# Use 'custom.law' argument to instruct the MortalityLaw function how to behave
my_model <- MortalityLaw(x = ages,
                         Dx = deaths, 
                         Ex = exposure, 
                         custom.law = my_gompertz)
summary(my_model) # these are the results
plot(my_model)
```





